Supplementary Materials for \"COATi: statistical pairwise alignment of protein coding sequences\

### 2023-12-28 alignpair_letter.tex
Check in revisions to methods.

The COATi FST has a large state space to keep track of codon substitution rates when codons might be interspersed with indel events. This additional state space increases the computational complexity of the alignment algorithm. To reduce the runtime complexity of COATi, we have also developed an approximation of the COATi FST that can be implemented with standard dynamic programming techniques. This approximation uses a marginal substitution model where the output nucleotides are independent of one another and only depend on the input codon and position. This produces a $\left(61 \times 3 \right) \times 4$ substitution model and eliminates the need to track dependencies between output nucleotides.

A marginal substitution model is calculated from a standard substitution model by calculating the marginal probabilities that each ancestral codon produces specific descendant nucleotides at each reading frame position. Specifically, let
%
\[
P_\text{cod}\left( Y_1 = y_1, Y_2 = y_2, Y_3 = y_3 |
                   X_1 = x_1, X_2 = x_2, X_3 = x_3 \right)
\]
%
represent transition probabilities from a codon model, and
%
\[
P_\text{mar}\left(Y_p = y \middle| X_1 = x_1, X_2 = x_2, X_3 = x_3 \right)
=
\sum_{y_1, y_2, y_3} I(y_p = y)
P_\text{cod}\left( y_1, y_2, y_3 |
                   x_1, x_2, x_3 \right)
\]
%
represent the marginal transition probabilities, where
$p \in \{1, 2, 3\}$ is the position of the descendant nucleotide relative to the ancestral reading frame and $I()$ is an indicator function.
COATi contains marginal models for both Muse and Gaut (\citeyear{muse_gaut_1994}) or the Empirical Codon Model, resulting in the marginal models codon-marginal-mg (default model) and codon-marginal-ecm.
These models emphasize the position in a codon where the substitution occurs, help restrict the effects of low-quality data in the descendant sequence, and allow more than one substitution per codon.
In combination with the indel model, alignment using the marginal model is implemented using dynamic programming.

@@

I downloaded 16000 human genes and their gorilla orthologs from the ENSEMBL
database \citep{ensembl_hubbard_2002}.
After downloading, I removed 2232 sequence-pairs longer than 6000 nucleotides and aligned the remaining pairs with all five methods.
At least one aligner added gaps to 6048 sequence pairs, and no aligner added gaps to 7761 sequence pairs.
Then, I randomly introduced gap patterns extracted from all five methods into the ungapped sequence pairs to generate the benchmark alignments.

The simulation algorithm can introduce a pairwise alignment pattern to any two nucleotide sequences of equal length. The alignment pattern is given as a CIGAR string (Compact Idiosyncratic Gapped Alignment Report), a format commonly used to summarize aligned reads to a reference genome. Assigning one of the sequences as the reference, to distinguish between insertions and deletions, CIGAR strings can also summarize pairwise alignments by grouping the number of contiguous matches or mismatches `M', deletions `D', and insertions `I'. The resulting pattern combines these letters preceded by the number of characters for each section as they appear in the alignment. This pattern is introduced by replacing nucleotides with gaps as indicated by deletions on one sequence and randomly introducing residues where the CIGAR strings indicated insertions.

@@

I created the benchmark of alignments by using an equal number of randomly sampled gap patterns from each aligner.
% After creating the dataset, I removed the gaps and measured how well different aligners were able to retrieve the true alignments.
I used the dataset to evaluate the accuracy of COATi and a suite of popular aligners spanning various alignment methods:
Clustal$\Omega$ v1.2.4 \citep{clustal_omega_sievers_2011},
MACSE v2.06 \citep{ranwez_macse_2011}, MAFFT v7.505
\citep{katoh2013mafft}, and PRANK v.150803 \citep{prank_loytynoja_2014}.

\subsection*{Metrics}

To quantify the similarity between each alignment in the benchmark and the corresponding output obtained by the different tools, I used the alignment error metric $d_{seq}$ \citep{metrics_blackburne_whelan_2011}. This metric accounts for indels and is more informative than conventional distance scores like sum-of-pairs or total columns. Intuitively, $d_{seq}$ ranges between zero and one and can be interpreted as the probability that a randomly selected residue will be aligned to a different location against a sequence that does not contain such residue.

The computation of $d_{seq}$ involves characterizing the gaps present in the alignment. Then, a site-wise homology set $H(A)^i_j$ is calculated for each alignment $A$, sequence $i$, and character $j$. The distance between two alignments $A$ and $B$ is the average across all characters of the symmetric difference (or Hamming distance), represented as `$\triangle$' between homology sets over the length of such sets:

\begin{equation}
d(A,B) = \frac{1}{c} \sum_i \sum_j \frac{|H(A)^i_j \triangle H(B)^i_j|}{|H(A)^i_j|+|H(B)^i_j|}
\end{equation}

\noindent where $c$ is the sum of the sequence lengths.

In addition, I compared the number of perfectly and imperfectly retrieved alignments for each aligner. Perfect alignments are defined as those with a distance of zero to the reference alignment ($d_{seq} = 0$), indicating 100\% similarity. Notably, a set of sequences can have more than one optimal alignment under the same evolutionary model (same score), despite algorithms typically producing a single result. Consequently, to account for evolutionary equivalent alignments, I scored all alignments using the marginal model and also considered perfect those with scores identical to the benchmark.
Furthermore, I counted the number of alignments with the lowest distance $d_{seq}$ to the true alignment, including ties, reported as best alignments. Moreover, I computed the count of imperfect alignments, where an alignment is considered imperfect when its distance to the reference alignment is greater than zero ($d_{seq} > 0$) and another method successfully produced an alignment with 100\% similarity. This analysis exposes instances where all aligners fall short of achieving a perfect result in addition to a direct comparison.

@@

To evaluate how well the aligners were able to identify positive and negative selection, I estimated $k_s$ and $k_a$ statistics. $k_s$ and $k_a$ are, respectively, the number of substitutions per synonymous site (no changes at the amino acid level) and per non-synonymous site (introduces changes at the amino acid level) between two protein-coding genes. They are also denoted as $d_s$ and $d_n$ in the literature. I used the R package seqinr \citep{seqinr} to estimate these metrics, which follows the popular method put forth by Li \citep{ka_ks_li_1993}. First, this method takes two aligned homologous protein-coding sequences and classifies the nucleotide sites in a sequence as nondegenerate, twofold degenerate, and fourfold degenerate. A site is nondegenerate if all possible changes at that site are nonsynonymous, twofold degenerate if one of the three possible changes is synonymous, and fourfold degenerate if all possible changes are synonymous. Second, the nucleotide changes between the two sequences are counted and divided as transitional (A$\leftrightarrow$G, C$\leftrightarrow$T) and transversional (\{A, G\}$\leftrightarrow$\{C, T\}). Third, the Kimura two-parameter distance \citep{kimura1980simple} is used to estimate the number of transitions and transversions per site type (nondegenerate, twofold degenerate, and fourfold degenerate), which is used as a correction factor for multiple hits. Finally, $k_s$ is the estimate of the average transitional rate at twofold and fourfold degenerate sites, and $k_a$ is the estimate of the average transversional rate at nondegenerate and twofold sites. In the results, these metrics are reported as the F$_1$ score, which is the harmonic mean of precision (true positives over total positives) and recall (true positives over true positives and false negatives). This score ranges between 0 and 1, with a score of 1 representing a perfect result.

In addition, we compared the estimated evolutionary distance between the reference and inferred alignments using the Kimura 2-parameter model \citep{kimura1980simple}. The calculation of the Kimura 2-parameter model involves considering the rates of nucleotide transitions and transversions. The formula used for distance calculation is $D = -0.5 \cdot \log((1 - 2P - Q) \cdot \sqrt{1 - 2Q})$, where $P$ represents the proportion of transitional substitutions and $Q$ represents the proportion of transversional substitutions. The resulting distance provides a quantitative measure of the evolutionary divergence between the sequences. These calculations were performed using the R software package ape \citep{paradis2019ape}.

>>>>

The COATi FST has a large state space to keep track of codon substitution rates when codons can be interspersed with indel events. This additional state space increases the computational complexity of the alignment algorithm. To reduce the runtime complexity of COATi, we have also developed an approximation of the COATi FST that can be implemented with standard dynamic programming techniques. This approximation uses a marginal substitution model where the output nucleotides are independent of one another and only depend on the input codon and position. This produces a $\left(61 \times 3 \right) \times 4$ substitution model and eliminates the need to track dependencies between output nucleotides.

A marginal substitution model is calculated from a standard substitution model by calculating the marginal probabilities that each ancestral codon produces specific descendant nucleotides at each reading frame position. Specifically, let
%
\[
P_\text{cod}\left( Y_1 = y_1, Y_2 = y_2, Y_3 = y_3 |
                   X_1 = x_1, X_2 = x_2, X_3 = x_3 \right)
\]
%
represent transition probabilities from a codon model, and
%
\[
P_\text{mar}\left(Y_p = y \middle| X_1 = x_1, X_2 = x_2, X_3 = x_3 \right)
=
\sum_{y_1, y_2, y_3} I(y_p = y)
P_\text{cod}\left( y_1, y_2, y_3 |
                   x_1, x_2, x_3 \right)
\]
%
represent the marginal transition probabilities, where $p \in \{1, 2, 3\}$ is the position of the descendant nucleotide relative to the ancestral reading frame and $I()$ is an indicator function. COATi contains marginal models for both Muse and Gaut (\citeyear{muse_gaut_1994}) or the Empirical Codon Model, resulting in the marginal models codon-marginal-mg and codon-marginal-ecm. These models emphasize the position in a codon where the substitution occurs, help restrict the effects of low-quality data in the descendant sequence, and allow more than one substitution per codon. In combination with the indel model, alignment using the marginal model is implemented using dynamic programming.

@@

In order to compare COATi against other aligners on realistic datasets, we developed a procedure to introduce realistic gap patterns into human-gorilla orthologous gene pairs that did not previous contain indels. Briefly, we downloaded 16000 human genes and their gorilla orthologs from the ENSEMBL
database \citep{ensembl_hubbard_2002}. After downloading, we removed 2232 sequence-pairs longer than 6000 nucleotides and aligned the remaining pairs with five alignment methods: COATi, Clustal$\Omega$ v1.2.4 \citep{clustal_omega_sievers_2011}, MACSE v2.06 \citep{ranwez_macse_2011}, MAFFT v7.505
\citep{katoh2013mafft}, and PRANK v.150803 \citep{prank_loytynoja_2014}. Next we separated sequence pairs into two sets: (1) 6048 sequence pairs for which at least one aligner added gaps, and (2) 7761 sequence pairs for which all five aligners did not introduce any gaps.

We identified gap patterns from the sequence pairs that were aligned with gaps and randomly introduced them into the ungapped sequence pairs. We used an equal number of randomly sampled gap patterns from each aligner to produce a benchmark set of known alignments. The phases of gaps were preserved as were the spacings of any clusters of gaps. Segments of matches that were 99 nucleotides or longer were allowed to change length to accommodate the length of the ungapped sequence pair. (This criteria was recursively lowered if gap pattern did not fit the ungapped sequence pair.) We deleted nucleotides from the gorilla sequences of ungapped pairs to create deletions, and we inserted random nucleotides according to their stationary frequency to the gorilla sequences to create insertions. The human sequences were unchanged.

% The simulation algorithm can introduce a pairwise alignment pattern to any two nucleotide sequences of equal length. The alignment pattern is given as a CIGAR string (Compact Idiosyncratic Gapped Alignment Report), a format commonly used to summarize aligned reads to a reference genome. Assigning one of the sequences as the reference, to distinguish between insertions and deletions, CIGAR strings can also summarize pairwise alignments by grouping the number of contiguous matches or mismatches `M', deletions `D', and insertions `I'. The resulting pattern combines these letters preceded by the number of characters for each section as they appear in the alignment. This pattern is introduced by replacing nucleotides with gaps as indicated by deletions on one sequence and randomly introducing residues where the CIGAR strings indicated insertions.

@@

% I created the benchmark of alignments by using an equal number of randomly sampled gap patterns from each aligner.
% % After creating the dataset, I removed the gaps and measured how well different aligners were able to retrieve the true alignments.
% I used the dataset to evaluate the accuracy of COATi and a suite of popular aligners spanning various alignment methods:
% Clustal$\Omega$ v1.2.4 \citep{clustal_omega_sievers_2011},
% MACSE v2.06 \citep{ranwez_macse_2011}, MAFFT v7.505
% \citep{katoh2013mafft}, and PRANK v.150803 \citep{prank_loytynoja_2014}.

\subsection*{Alignment Accuracy Measures}

We used multiple statistics to to quantify the similarity between each alignment in the benchmark and the corresponding output obtained by the different tools.

\paragraph{Alignment Error.}
To quantify the error between estimated alignments and the benchmark alignments, we used the alignment error metric $d_{seq}$ \citep{metrics_blackburne_whelan_2011}. Intuitively, $d_{seq}$ ranges between zero and one and can be interpreted as the fraction of nucleotides in the sequence pair that are aligned differently between estimated and benchmark alignments. This metric summarizes each alignment by building homology sets for each nucleotide in the sequence pair. Briefly, if an alignment column contains position $i$ from the first sequence and position $j$ from the second sequence, then $\{j\}$ is the homology set for position $i$ in the first sequence and $\{i\}$ is the homology set for position $j$ in the second sequence. If a position is aligned against a gap, then its homology set is simply $\{\text{gap}\}$. This treats all gaps equally as the location of the gap is not recorded. Finally, $d_{seq}$ is calculated  between an estimated and a benchmark alignment as the average, normalized Hamming distance between the homology sets for each nucleotide in the sequence pair. See \cite{metrics_blackburne_whelan_2011} for further details.

% The computation of $d_{seq}$ involves characterizing the gaps present in the alignment. Then, a site-wise homology set $H(A)^i_j$ is calculated for each alignment $A$, sequence $i$, and character $j$. The distance between two alignments $A$ and $B$ is the average across all characters of the symmetric difference (or Hamming distance), represented as `$\triangle$' between homology sets over the length of such sets:

% \begin{equation}
% d(A,B) = \frac{1}{c} \sum_i \sum_j \frac{|H(A)^i_j \triangle H(B)^i_j|}{|H(A)^i_j|+|H(B)^i_j|}
% \end{equation}

% \noindent where $c$ is the sum of the sequence lengths.

\paragraph{Perfect, Imperfect, and Best Alignments.}
We quantified the number of perfect, imperfect and best alignments each aligner produced. We define perfect alignments as alignments with a distance of zero to their benchmark alignment ($d_{seq} = 0$) or any alignment that is evolutionary equivalent to an alignment with a distance of zero. Equivalency was included in our definition of perfect alignments to prevent the manner by which aligners break ties between multiple optimal alignments from impacting our results. Evolutionary equivalence was determined by scoring alignments using COATi's marginal model. Any alignment that had a score which matched the score of its benchmark alignment was considered perfect even if its distance to the benchmark was greater than zero.
%
Imperfect alignments are defined as alignments that are not perfect when another method successfully produces a perfect alignment for the same pair of sequences.
%
Best alignments are the alignments that the lowest distance $d_{seq}$ to the true alignment, including ties.
%
Taken together, these three statistics not only allow a direct comparison of aligners but also expose instances where all aligners fall short of achieving a perfect result.

@@

% Ks and Ka are, respectively, the number of substitutions per synonymous site and per non-synonymous site between two protein-coding genes.

\paragraph{Selection.}
Alignment error negatively impacts the identification of genes that have experienced positive and negative selection. To evaluate the impact of aligners on selection identification, we estimated $k_s$ and $k_a$ statistics (also known as $d_s$ and $d_n$) for each estimated alignment and compared them to the benchmark alignment. Briefly, $k_s$ is the number of substitutions per synonymous site and $k_a$ is the number of substitutions per non-synonymous site between two protein-coding sequences. We used the method of \cite{ka_ks_li_1993} to estimate these statistics as implemented in the R package seqinr \citep{seqinr}. 
%
We considered any alignment to be showing evidence of positive selection if $k_a/k_s > 1$ and negative selection if $k_a/k_s < 1$. We estimated F\textsubscript{1} scores for both positive and negative selection by comparing estimated alignments to benchmark alignments. F\textsubscript{1} is the harmonic mean of precision and recall and allows us to measure how well aligners produce alignments that correctly identify or exclude positive selection or negative selection.

% First, this method takes two aligned homologous protein-coding sequences and classifies the nucleotide sites in a sequence as nondegenerate, twofold degenerate, and fourfold degenerate. A site is nondegenerate if all possible changes at that site are nonsynonymous, twofold degenerate if one of the three possible changes is synonymous, and fourfold degenerate if all possible changes are synonymous. Second, the nucleotide changes between the two sequences are counted and divided as transitional (A$\leftrightarrow$G, C$\leftrightarrow$T) and transversional (\{A, G\}$\leftrightarrow$\{C, T\}). Third, the Kimura two-parameter distance \citep{kimura1980simple} is used to estimate the number of transitions and transversions per site type (nondegenerate, twofold degenerate, and fourfold degenerate), which is used as a correction factor for multiple hits. Finally, $k_s$ is the estimate of the average transitional rate at twofold and fourfold degenerate sites, and $k_a$ is the estimate of the average transversional rate at nondegenerate and twofold sites. In the results, these metrics are reported as the F$_1$ score, which is the harmonic mean of precision (true positives over total positives) and recall (true positives over true positives and false negatives). This score ranges between 0 and 1, with a score of 1 representing a perfect result.

\paragraph{Evolutionary Distance.}
Alignment error also negatively impacts the estimation of evolutionary distances and phylogenetic trees. To quantify the impact of aligners on the estimation of evolutionary distances, we compared Kimura's 2-parameter distance \citep[K2P;][]{kimura1980simple} calculated for the estimated and benchmark alignments. This distance corrects for multiple mutations at a site and takes into account differences in transition and transversion rates. It also assumes equal nucleotide frequencies and no variations in the rate of substitution across sites. While Kimura's 2-parameter distance is more suitable for non-coding sequences, it is straight forward to calculate and provides a quantitative measure of the evolutionary divergence between sequences. We used the R software package ape \citep{paradis2019ape} to calculate K2P distances.

% e compared the estimated evolutionary distance between the reference and inferred alignments using the Kimura's 2-parameter distance \citep{kimura1980simple}. The calculation of the Kimura 2-parameter model involves considering the rates of nucleotide transitions and transversions. The formula used for distance calculation is $D = -0.5 \cdot \log((1 - 2P - Q) \cdot \sqrt{1 - 2Q})$, where $P$ represents the proportion of transitional substitutions and $Q$ represents the proportion of transversional substitutions.
